Towards an Ecological Trait-data Standard

### 2017-11-28 manuscript/manuscript.Rmd
update review section: merge ontology concepts from later sections, remove data table formats (for now)

# 1. Need for data standardisation

@@

# 2. A review of trait-data initiatives

@@

In the field of comparative physiology, morphological traits such as plant flower, leaf and stem traits or bird wing and beak measurements, as well as life-history traits such as Ellenberg values  have been measured for decades, and have been contained in regular articles or books. With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that comprise a local set of taxa and a focal set of traits. 
A plethora of such static datasets has been published along with scientific articles [see @kleyer08 for a review on plant data; on animal data, see e.g., @sokol11; @gossner15; @gossner15c; @renner14; @vanderplas17; @price13].

@@

## Trait databases

<!-- community database backend, but static publication --> 
In the past two decades, many distributed datasets have been aggregated and harmonized into greater collections with particular taxonomic or regional focus [e.g. @klotz02; @iversen17; @kleyer08; @ricklefs17; @oliveira17; @iversen17; @myhrvold15; @kissling14; @jones09]. [@wright04; ]
Most of these data are being published as static datasets and are being updated irregularly. Documentation of metadata and methodology differs in detail, depending on the research focus of the initiative. 
Few of these datasets are using a standardised terminology for taxa or for traits that would allow to relate them to other databases.

<!-- database backend, selective download, user access management --> 
A more dynmically growing, advanced online portal,  offer 
selective data download 
user access and rights management 
webservices (API)

most prominently, plant traits from many different sources have been collated into the extensive TRY database [@kattge11].  Traits on crop plants have been collected in agronomy databases like Planteome and BetyDB [https://www.betydb.org/; @lebauer17].

Specialised trait-databases have been created that cover different taxa, interaction types (e.g. pollination traits or feeding relationships), or ecosystem types. For example, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).

@@

Unlike for plants, a single unified webservice and harmonizing scheme for animal trait data is still lacking. The reason for this may be that data on animals, which span multiple trophic levels and possess diverse body plans make harmonizing animal trait data a more complex task than for plant traits [@moretti17].

On the other hand, not all centralized databases for functional traits comply with the criteria of open access, which leaves them at a disadvantage compared to low-threshold repositories.

@@

In database management, a glossary or thesaurus of terms is primarily a listing of definitions. Expanding on this, ontologies link the defined terms with each other by formally defining the relationships between them, with the objective to enable a computational interpretation of data. 
Relating the trait definition to methodological definitions on trait measurements is key for harmonizing trait data already at the physical process of measuring. For instance, handbooks for measurement protocols provide such a methodological standardization for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17] and should be referenced when defining traits in thesauri and ontologies.

In the context of trait-based research, several thesauri of trait definitions have been developed for plants, for instance in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], the Planteome ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org], or the Flora Phenotype Ontology [http://obofoundry.org/ontology/flopo.html, @hoehndorf16]. On the side of animal traits, a couple of initiatives published comprehensive trait ontologies, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14], the Hymenoptera Anatomy Ontology [@yoder10, http://hymao.org/], or the vertebrate trait ontology [http://obofoundry.org/ontology/vt.html, @park13]. 
More general classes of traits are defined within the GeoBON Essential Biodiversity Variables [@pereira13].
Platforms hosting biological ontologies and terminologies are the OBO foundry (http://obofoundry.org/) or ontobee (http://www.ontobee.org/).

Most of these ontologies are rather simple in structure, and mainly concerned with defining terms, rather than defining relationships of terms. The degree to which they define body-parts, life-history properties, and measurement procedures, or how they define expected standard units, differs widely.

## Ontologies for the semantic web

A general database for major physiological and life-history traits of all kingdoms is the Encyclopedia of Life Trait Bank project [@parr16, see below for more detail].

While trait thesauri define traits for focal groups of organisms, they are not specifying the  format or structure in which trait data should be stored. 
Recently, the Encyclopedia of Life (EOL) project has advanced on the topic by developing the database framework TraitBank [@parr16], which is to date the most general approach to providing a semantic structure for trait data tables. The framework employs established terms provided by the Darwin Core Standard [DWC, @wieczorek12], relates trait definitions to specific trait, phenotypic and anatomical ontologies, and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g. GBIF Taxonomy) to capture synonyms, misspellings and controversies [@parr16]. Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive. 
Another initiative aimed for a mobilization of measurements from museal collection data collated in the VertNet database into harmonized traitdata [@guralnick16].

However, the Terms of the Darwin Core Standard and available Extensions, TraitBank and VertNet, are responses to a demand in biodiversity data management for structured input from the research communities. 
The practical importance of these frameworks for data providers and data users is less obvious and often not appropriately considered when storing or uploading data to low-threshold  file-publishing services. None of the aggregated and harmonized databases for birds, amphibians, or mammals employs such ontologies or semantic web standards. One reason certainly is compatibility: to date, the existing trait data standards do not cover all of the aspects required in the various use-cases of trait-based research. <!-- examples for blind spots --> 
The other reason is competence: trait-data providers are not well trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without immediate pay-off.

# 3. The structure of trait datasets

As a minimum consensus, trait datasets may be defined as follows: 
a trait dataset contains quantitative measurements or qualitative facts (i.e. trait values) about physical phenotypic characteristics of fitness, behaviour or performance (i.e. traits) observed on an individual organism or parts of an individual (i.e. an entity) that can be assigned to a biological taxon (i.e. a species or higher-level taxon) <!-- or larger organisational unit, e.g. a community -->.

Quantitative measurements are values obtained either by direct morphological or physiological measurement on single specimens (Fig. 1*a*), by aggregation of replicated measurements on multiple entities (Fig. 1*b*) or by estimation of means or ranges for the species as reported in literature or other published sources (e.g. databases, Fig. 1*c*). Qualitative facts are assignments of an entity to a categorical level, e.g. of a behavioural or life-history trait (Fig. 1.*d*). The entity or observation (i.e. the occurrence) to which the reported measurement or fact applies may differ in resolution -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or higher-level taxon.  
This definition refines the entity-quality model (EQ) as described by @garnier17, where a trait observation is phrased as 'an entity having a quality'. 
<!-- Cat: the difference [between quantitative measurements or qualitative facts] is not totally clear to me - is it the difference between field vs literature gathered data?  Flo: I see measurements as quantitative and facts as qualitative traits, which mostly aligns with field vs. literature, but not entirely. A differentiation that has rarely been specified in the trait-litereature it seems. Garnier et al 2017 just speak of "quality", which is assigned to an "entity", without specifiying it further. Nadja: I like this differentiation, to me this nicely reflects morphometric and literature traits (although its of course not exclusive). Maybe it helps if we add an example after the definition? --> <!--FDS: is this sufficiently defined now? the distinction between measured and literature data is spelled-out further below, when discussing the formats. Quantitative values can be taken from literature, though, like body sizes for example. -->

@@

Trait-datasets usually take different formats. For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species$\times$trait matrix (a.k.a. wide-table format,  Fig. 2*a*), with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available. This format is usually reporting missing data as 'NA'. It may store additional information (e.g. on variation of means or literature source) in secondary columns or accompanying datasheets. 
The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges for this taxon). Quantitative traits at this resolution for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10a]. Also at this level of information, missing entries, e.g. about the trait values or behaviour and functional role of a rarely described species, may be inferred from these tables combined with phylogenies ['gap filling', @swenson14] and using Bayesian approaches [@schrodt15]. However, the tables do not allow to estimate within-species variation or assess the quality of the reported measurement or fact [@guralnick16].

@@

@@

# 4. Proposing an ecological traitdata standard

With this paper, we propose a versatile, minimal data structure, as well as a defined vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based research. 
We were drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).  
We were particularly considering the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of datasets with major web-services and existing standards for biodiversity data management.

@@

In very complex datasets repetition of verbose entries and long character strings in a single data table should be avoided [@wickham14]. Therefore we recommend use of  Darwin Core Archives [DwC-A, @robertson09], the structure proposed for GBIF [@gbif17,  http://tools.gbif.org/dwca-assistant/] or EOL TraitBank [@parr16,  http://eol.org/info/cp_archives]. This also integrates a defined framework for metadata in the Ecological Metadata Language [@knb11], which should be used for describing and linking the data tables.

@@

It is indispensable to relate the user-defined vocabulary to a consensus standard of taxon names or or look-up tables of traits. This is achieved by adding globally valid identifiers for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the more readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively), according to a public ontology (e.g. GBIF Backbone Terminology; for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass').  
By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation and may refer to different scientific community standards  [as practised in TRY; @kattge11]. This redundancy of data enables continuity on the author and data owner side and quality checking and comparability on the data user side. Keeping an additional reference to a kingdom and class (using columns `kingdom`, `class` as described by DWC and suggested for EOL Traitbank) will avoid ambiguities with taxonomic homonyms. 
It is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level. This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorisation [@schrodt15].

The existing thesauri and ontologies for traits differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control. The accessibility by software differs widely. For biodiversity data, the OBO Foundry (http://www.obofoundry.org/) or the GFBio Terminology service (https://terminologies.gfbio.org/browse/), provide hosting for advanced trait ontologies and offer API services for computational access. More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.

If no published trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus. A thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17]. 
Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g. units or legit factor levels) of measured values or reported facts. It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits. 
By providing a minimal vocabulary for this case (https://ecologicaltraitdata.github.io/ETS/#terms-for-traitlists-a-trait-thesaurus), we hope to facilitate the publication of trait thesauri developed for the own project context which always should be referenced in the core trait dataset. The thesaurus might accompany the core data in a Darwin Core Archive, or be published on any other stable webservice.

@@

Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package offers direct access to trait datasets that have been published in the public domain or under creative commons licenses [@gossner15; @gossner15c; @ricklefs17; @vanderplas17; @myhrvold15]. See a full list of linked datasets on the package  website or by calling `data(package = "traitdataform")`. We invite users and authors of datasets to add further data to the package and thereby contribute to this registry for distributed trait datasets.

>>>>

@@

# 1. A review of trait-data initiatives

@@

In the field of comparative physiology, morphological traits such as plant flower, leaf and stem traits or bird wing and beak measurements, as well as life-history traits such as Ellenberg values  have been measured for decades, and have been contained in regular articles or books. With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that comprise a local set of taxa and a focal set of traits. 
A plethora of such static datasets has been published along with scientific articles [see @kleyer08 for a review on plant data; on animal data, see e.g., @sokol11; @gossner15; @gossner15c; @renner14; @vanderplas17; @price13;  @ricklefs17].

@@

## Database initiatives

<!-- community database backend, but static publication --> 
In the past two decades, many distributed datasets have been aggregated and harmonized into greater collections with particular taxonomic or regional focus [e.g. @klotz02; @iversen17; @kleyer08; @oliveira17; @iversen17; @myhrvold15; @kissling14; @jones09]. <!--[@wright04; ] -->
Another initiative aimed for a mobilization of measurements from museal collection data collated in the VertNet database into harmonized traitdata [@guralnick16]. 
Most of these databases are published as static datasets and are updated irregularly. Documentation of metadata and methodology differs in detail, depending on the research focus of the initiative. 
Few of these datasets are using a standardised terminology for taxa or for traits that would allow to relate them to other databases.

<!-- database backend, selective download, user access management --> 
More specialised trait-databases have been created that cover different taxa, interaction types (e.g. pollination traits or feeding relationships), or ecosystem types. These centralized database initiatives invite data submissions from  a defined research field and offer to take care of the harmonization process and greatly facilitate data synthesis. 
For example, traits on crop plants have been collected in agronomy databases like Planteome and BetyDB [https://www.betydb.org/; @lebauer17].
The BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).

By aiming for a universal framework for plant traits, the TRY database [@kattge11]  attracted more data submissions and downloads than any other platform. The online platform enables selective data download and user permission and rights management. Being a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section). 
Unlike for plants, a single unified webservice and harmonizing scheme for animal trait data is still lacking. The reason for this may be that data on animals, which span multiple trophic levels and possess diverse body plans make harmonizing animal trait data a more complex task than for plant traits [@moretti17].

@@

Regarding open access, not all centralized databases for functional traits comply with the criteria demanded by journals and funding agencies, which leaves them at a disadvantage compared to low-threshold repositories.

@@

In database management, a glossary or thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17].  Expanding on this, ontologies link the defined terms with each other by formally defining the relationships between them, with the objective to enable a computational interpretation of data. Providing relational ontologies in a machine readable form makes them compliant with semantic web standards, thus supporting the goal of widely interlinked data structures for the world wide web [@berners-lee01; @page08].

<!--Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g. units or legit factor levels) of measured values or reported facts. It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits. -->

Standard trait definitions have been provided in all shades along this continuum, ranging from listings of definitions in glossaries, over well-defined thesauri to relational ontologies. The initiatives behind these are first and foremost concerned with consensus building on trait definitions. 
Assigning an unambiguous methodological protocol of measurements to a trait is key for harmonizing the physical process of measuring. For instance, handbooks for measurement protocols provide such a methodological standardization for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17] and should be referenced when defining traits in thesauri and ontologies.
Very general classes of traits are defined within the list of GeoBON Essential Biodiversity Variables [@pereira13].
Thesauri of trait definitions have been developed for plants [e.g. the Planteome ontology for crop plants, @walls12; @jaiswal05; the Flora Phenotype Ontology, @hoehndorf16], as well as for animals [e.g. the Hymenoptera Anatomy Ontology, @yoder10; the vertebrate trait ontology, @park13]. 
Comprehensive trait ontologies, including hierarchies of parent terms, have been developed in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], and on the side of animal traits, in the Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14].

These existing thesauri and ontologies for traits differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control [@]. 
For biodiversity data, the OBO Foundry (http://www.obofoundry.org/), ontobee (http://www.ontobee.org/), or the GFBio Terminology service (https://terminologies.gfbio.org/browse/), provide hosting for advanced trait ontologies and offer webservices for computational access. 
These publicly available online resources also enable linking trait measurements to globally unique Uniform Resource Identifiers (URIs) when referencing traits from within own datasets. For example, a measurement of seed size would be linked to Planteome Trait Ontology definition of 'seed size' by referencing 'http://browser.planteome.org/amigo/term/TO:0000391'.

To be unambiguous, any thesaurus should be defining terms based on other well-defined terms from semantic ontologies. The definition provided here contains references to other globally defined terms: "A seed morphology trait (TO:0000184) which is the size of a seed (PO:0009010)."
This way, each trait definition may link to a broader or narrower term. For example, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait'. 
Furthermore, trait definitions may refer to related terms or synonyms defined in other trait ontologies. By providing this semantic interlinkage of trait ontologies, a web of definitions is spun across the internet which allows researchers and search engines to relate independent trait measurements with each other.

The benefit of such ontologies will increase if webservices and open Application Programming Interfaces (APIs) provide a way to extract the definitions and higher-level trait hierarchies programmatically via software tools. To harmonize trait data across databases, future trait standard initiatives should provide this functionality.
More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparitive study of trait data from different sources.

## Trait-data structures for the semantic web

While trait thesauri define traits for focal groups of organisms, they are not specifying the  format or structure in which trait data should be stored and linked to further standard terminologies, e.g. for taxa. 
To make sense of trait data in the context of more general databases, a consensus definition of trait datasets is inevitable.

Trait data have been defined  by @garnier17 to follow an entity-quality model (EQ), where a trait observation is phrased as 'an entity having a quality'. 
More specifically, a trait dataset contains quantitative measurements or qualitative facts (i.e. trait values) about physical phenotypic characteristics of fitness, behaviour or performance (i.e. traits) observed on an biological entity (i.e. an individual specimen, or parts of an individual) that can be assigned to a biological taxon (i.e. a species or higher-level taxon). 
Expaning on this definition, quantitative measurements are values obtained either by direct morphological or physiological measurement on single specimens (Fig. 1*a*), by aggregation of replicated measurements on multiple entities (Fig. 1*b*) or by estimation of means or ranges for the species as reported in literature or other published sources (e.g. databases, Fig. 1*c*). Qualitative facts are assignments of an entity to a categorical level, e.g. of a behavioural or life-history trait (Fig. 1.*d*). 
The entity or observation (i.e. the occurrence) to which the reported measurement or fact applies may differ in resolution -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or higher-level taxon.

@@

These relations between a trait observation and an individual specimen as an occurrence of a particular taxon have been formalized in the schema for biological collection records [ABCD Schema; @walls14] and the Darwin Core Standard for biodiversity data [DwC; @wieczorek12]. These frameworks specify terms and classes to describe the general structure of biodiversity databases, for example by defining which field is to contain taxon names, sex or life stage, time and date of observation, as well as measurement values and methodological detail.

Recently, the Encyclopedia of Life (EOL) project has proposed the database framework TraitBank [@parr16] for major physiological and life-history traits of all kingdoms, which is to date the most general approach to providing an integrated structure for trait data tables.
The framework employs established terms provided by the DwC, relates trait definitions to specific trait, phenotypic and anatomical ontologies, and maps taxa to global identifiers in taxonomic hierarchies of name service providers  to capture synonyms, misspellings and controversies [@parr16, http://eol.org/info/cp_archives]. For example, the Common Daisy (*Bellis perennis*) would be related via a globally unique taxonID using the URI https://www.gbif.org/species/3117424 from the GBIF Backbone Terminology. 
Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. 
TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive [DwC-A, @robertson09], a zip-file with annotated text-files that is also preffered for observation data in GBIF [@gbif17,  http://tools.gbif.org/dwca-assistant/]. 
The archive also integrates a defined framework for metadata in the Ecological Metadata Language [@knb11], which should be used for describing and linking the data tables.

However, the terms of the Darwin Core Standard and available extensions, as well as TraitBank, are responses to a demand in biodiversity data management for structured input from the research communities.

The practical importance of these frameworks for data providers and data users is less obvious and often not appropriately considered when storing or uploading data to low-threshold  file-publishing services. None of the aggregated and harmonized databases for birds, amphibians, or mammals employs such ontologies or semantic web standards. One reason certainly is compatibility: to date, the existing trait data standards do not cover all of the aspects required in the various use-cases of trait-based research. <!-- examples for blind spots --> 
The other reason is competence: trait-data providers are not well trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without immediate pay-off.

<!--
Depending on the research objective, trait-data are stored in a range of different table formats. In the subsequent paragraphs, we discuss the most common formats  in relation to the use cases in trait-based ecology.

### a) Species wide-table

If trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species$\times$trait matrix (Fig. 2*a*), with a column of values for each trait recorded and a row for each species (or taxon) for which data were available. 
This format needs to clearly differentiate missing data and traits not applicable to a given taxon. It may store additional information (e.g. on variation of means or literature source) in secondary columns or accompanying datasheets. 
The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges of trait values for this taxon).
Quantitative traits reported at this resolution are often used to calculate community weighted means or functional diversity metrics at the community level [@laliberte10a]. Also at this level of information, missing entries, e.g. about the trait values of a rarely described species, may be inferred from these tables when combined with phylogenies ['gap filling', @swenson14; @penone14] and using Bayesian approaches [@schrodt15]. <!--However, these matrices do not allow estimates of within-species variation or assessments of the quality of the reported measurement or fact [@guralnick16].

### b) Occurrence wide-table

@@

### c.) measurement long-table

@@

# 2. Proposing an ecological traitdata standard

With this paper, we propose a versatile, minimal data structure, as well as a defined vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based ecological research. 
We were drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).  
We were particularly considering the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of datasets with major web-services and existing standards for biodiversity data management.

@@

In very complex datasets repetition of verbose entries and long character strings in a single data table should be avoided [@wickham14]. Therefore we recommend use of  Darwin Core Archives [DwC-A, @robertson09].

@@

It is indispensable to relate the user-defined vocabulary to a consensus standard of taxon names or or look-up tables of traits. This is achieved by adding globally valid identifiers for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the more readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively)<!--, according to a public ontology (e.g. GBIF Backbone Terminology; for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass')-->.  
By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation and may refer to different scientific community standards  [as practised in TRY; @kattge11]. This redundancy of data enables continuity on the author and data owner side and quality checking and comparability on the data user side. Keeping an additional reference to a kingdom and class (using columns `kingdom`, `class` as described by DWC and suggested for EOL Traitbank) will avoid ambiguities with taxonomic homonyms. 
It is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level. This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorisation [@schrodt15].

<!--The existing thesauri and ontologies for traits differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control. The accessibility by software differs widely. For biodiversity data, the OBO Foundry (http://www.obofoundry.org/) or the GFBio Terminology service (https://terminologies.gfbio.org/browse/), provide hosting for advanced trait ontologies and offer API services for computational access. More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources. -->

If no published trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus. <!-- A thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17]. 
Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g. units or legit factor levels) of measured values or reported facts. It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits. -->
By providing a minimal vocabulary for this case (https://ecologicaltraitdata.github.io/ETS/#terms-for-traitlists-a-trait-thesaurus), we hope to facilitate the publication of trait thesauri developed for the own project context which always should be referenced in the core trait dataset. The thesaurus might accompany the core data in a Darwin Core Archive, or be published on any other stable webservice.

@@

Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package offers direct access to trait datasets that have been published in the public domain or under creative commons licenses [@gossner15; @gossner15c; @vanderplas17; @myhrvold15]. See a full list of linked datasets on the package  website or by calling `data(package = "traitdataform")`. We invite users and authors of datasets to add further data to the package and thereby contribute to this registry for distributed trait datasets.
